---
title: "Practical Machine Learning Course Project"
author: "Mark Edosa"
date: "July 25, 2016"
output: html_document
---

#Introduction
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement-a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

Data Sources
The training data for this project are available here:
        https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv
The test data are available here:
        https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

Downloading the data and loading into R
```{r}
if (!file.exists("pml-training.csv")) {
        download.file("http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", destfile = "pml-training.csv")
}
if (!file.exists("pml-testing.csv")) {
        download.file("http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", destfile = "pml-testing.csv")
        
}

testing <- read.csv("pml-testing.csv", header= TRUE) # read into R
training <- read.csv("pml-training.csv", header= TRUE)
```

Data Partitioning

the training data is split into training and validation data while the testing data is of course the test sample or new data.

```{r}
set.seed(1) 
library(caret)
Intrain <- createDataPartition(y=training$classe, p=0.7, list=FALSE)
trainingData <- training[Intrain,]
validation <- training[-Intrain,]
dim(trainingData); dim(validation)# view the dimensions
```

Cleaning the data


It is necessary to remove some variables that have no variability at all (i.e. variable indicating. These variables are not useful when we want to construct a prediction model. if nzv is true the variable is deleted

```{r}                                                                         
Nzvals <- nearZeroVar(trainingData, saveMetrics=TRUE) # shows the near zero covariates
                                                                          
## creating a subset without NZV variables:
                                                                          
Nzvals <- names(trainingData) %in% c("new_window", "kurtosis_roll_belt", "kurtosis_picth_belt", "kurtosis_yaw_belt", "skewness_roll_belt", "skewness_roll_belt.1", "skewness_yaw_belt", "max_yaw_belt", "min_yaw_belt", "amplitude_yaw_belt", "avg_roll_arm", "stddev_roll_arm", "var_roll_arm", "avg_pitch_arm", "stddev_pitch_arm", "var_pitch_arm", "avg_yaw_arm","stddev_yaw_arm", "var_yaw_arm", "kurtosis_roll_arm", "kurtosis_picth_arm","kurtosis_yaw_arm", "skewness_roll_arm", "skewness_pitch_arm", "skewness_yaw_arm","max_roll_arm", "min_roll_arm", "min_pitch_arm", "amplitude_roll_arm", "amplitude_pitch_arm", "kurtosis_roll_dumbbell", "kurtosis_picth_dumbbell", "kurtosis_yaw_dumbbell", "skewness_roll_dumbbell", "skewness_pitch_dumbbell", "skewness_yaw_dumbbell", "max_yaw_dumbbell", "min_yaw_dumbbell", "amplitude_yaw_dumbbell", "kurtosis_roll_forearm", "kurtosis_picth_forearm", "kurtosis_yaw_forearm","skewness_roll_forearm", "skewness_pitch_forearm", "skewness_yaw_forearm", "max_roll_forearm","max_yaw_forearm", "min_roll_forearm", "min_yaw_forearm", "amplitude_roll_forearm","amplitude_yaw_forearm", "avg_roll_forearm", "stddev_roll_forearm", "var_roll_forearm", "avg_pitch_forearm", "stddev_pitch_forearm", "var_pitch_forearm", "avg_yaw_forearm","stddev_yaw_forearm", "var_yaw_forearm")
                                                                          
newtData <- trainingData[!Nzvals]
dim(newtData)
newtData <- newtData[c(-1)] # romoving the ID
```                                                                          

Cleaning Variables with too many NAs here variables that have more than a 40% of NA's are removed

```{r}                                                                          
subsetnoNa <- newtData #creating another subset to iterate in loop
for(i in 1:length(newtData)) { # entire columns                                      
        if(sum(is.na(newtData[, i] ) ) /nrow(newtData) >= 0.4 ) { #if n?? NAs >       40% of total observations
        for(j in 1:length(subsetnoNa)) {
        if( length( grep(names(newtData[i]), names(subsetnoNa)[j]) ) ==1)  { #if the columns are the same:
         subsetnoNa <- subsetnoNa[ , -j] #Remove that column
        }   
   } 
 }
}
                                                                          
dim(subsetnoNa)
                                                                          
newTData1 <- subsetnoNa # refined training data
```

clean the validation and testing data sets as done above.

```{r}                                                                          
clean1 <- colnames(newTData1)
clean2 <- colnames(newTData1[, -58]) #already with classe column removed
validData <- validation[clean1]
testing1 <- testing[clean2] # i changed the testing to testing cause a noticed # errors when i try to subset with same name or variable.
dim(validData) ; dim(testing1)
```

coerce the data into the same type.
                                                                          
```{r}

 for (i in 1:length(testing1) ) {
          for(j in 1:length(newTData1)) {
          if( length( grep(names(newTData1[i]), names(testing1)[j]) ) ==1)  {              
          class(testing1[j]) <- class(newTData1[i])
       }     
}      
}
testing1 <- rbind(newTData1[2, -58] , testing1)
testing1 <- testing1[-1,]
```

Training and predicting

# 1. Decision Tree

```{r}
library(rpart)
library(rpart.plot)
modFit1 <- rpart(classe ~ ., data=newTData1, method = "class") # produces an rpart object, which can be used to predict new data/test values
rpart.plot(modFit1)                                                     

# Prediction:
predT <- predict(modFit1, validData, type = "class")
confusionMatrix(predT, validData$classe)                                                                           
```

# 2. Random Forests

```{r}
library(randomForest)
modfitrf = randomForest(classe ~ ., newTData1)
# Predicting in-sample error:
predRF <- predict(modfitrf, validData, type = "class")
confusionMatrix(predRF, validData$classe)
```                                                                                          
Random Forests yielded better Results.

# Answers for the Assignment

```{r}
# For Random Forests we use the following formula, which yielded a much better prediction in in-sample:
prediction2 <- predict(modfitrf, testing1, type = "class")
print(as.data.frame(prediction2))
```
